ediblecity: an R package to model and estimate the benefits of urban agriculture

### 2022-11-18 paper_v01.Rmd
create final sections

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(ediblecity)
library(knitr)
library(sf)
library(purrr)
library(tidyr)

@@

@@

Therefore, decision-makers, who are responsible of leading the urban transitions to more sustainable and resilient cities, are orphan of evidence in how to implement urban agriculture to maximize its impact on sustainability. Yet, several studies provided some insights that can guide the implementation of urban agriculture. For instance, some models assist planners in rainwater harvesting [@Gittleman2017; @Lupia2017]. Broader, @Gomez-Villarino2021 developed guidelines to maximize ecosystem services through urban agriculture. And, as expected, many models have been developed to quantify food production and food security provided by urban agriculture [@Grafius2020; @Grewal2012; @Hsieh2017; @MacRae2010].

Hence, our goal is to provide a model to estimate multiple benefits of urban agriculture that help decision-makers to strategically implement urban agriculture solutions. We developed the estimations for eight indicators and a functionality to create scenarios of urban agriculture based on the proportion of elements to be transformed to urban agriculture and on which elements will be transformed. In the methods, we present the main functionalities of the `ediblecity` package and then we apply those to an example in a neighborhood of Girona (Spain).

@@

## Urban representation of the city of interest

@@

city_functions |>
  select(-starts_with("...")) |> 
  select(-edible, -public, -jobs, -volunteers, -location) |>
  select(-starts_with("water")) |> 
  kable()
```
## Indicators estimated

### Urban heat island

@@

### Runoff prevention

@@

### Green areas accessibility

@@

### Nitrogen dioxide sequestration

@@

### Jobs created and volunteers involved

@@

### Green per capita

@@

### Food production

@@

## Scenarios of urban agriculture

@@

@@

To our understanding, the most important use case of the `ediblecity` package is to compare scenarios. To illustrate this, we will create two scenarios and compare them with the original `city_example`. The package has been designed to work well with the `tidyverse` framework in `R` [@Wickham2019], especially, with the `map_*` family in the `purrr` package [@Henry2022]. Therefore, we first create the two scenarios and save them in a list along with the original urban representation, which we called `s0` (`r ref("Code snippet", "use_set_scenario")`). The scenario `s1` will convert the 50% of elements to urban agriculture solutions while the scenario `s2` will convert the 100%. Half of gardens in vacant plots, streets and rooftop will have commercial purposes.

@@

# Create new scenarios
scenarios <- map(c(0.5, 1), ~set_scenario(city_example, 
                             pGardens = .x,
                             pVacant = .x,
                             pRooftop = .x,
                             private_gardens_from = "Normal garden",
                             vacant_from = c("Vacant", "Streets"),
                             rooftop_from = "Rooftop",
                             pCommercial = 0.5))

@@

# Name the scenarios
names(scenarios) <- c("s1", "s2", "s0")
```

@@

counting(scenarios[[3]], "s0") |> 
  full_join(counting(scenarios[[1]], "s1")) |> 
  full_join(counting(scenarios[[2]], "s2")) |> 
  mutate(across(where(is.numeric), ~if_else(is.na(.x), 0L, .x))) |> 
  kable()
```

## Calculate indicators and compare scenarios

@@

# We use the SVF object that is provided for the city_example in the package
map_dfr(scenarios, UHI, SVF = ediblecity::SVF) |> 
  mutate(Scenario = names(scenarios), .before = 1) |> 
  kable()

@@

As expected the second scenario (`s2`) has the lowest values for urban heat island, but no too far from the scenario `s1`. Both present a reduction of approximated 50% in urban heat island regarding the base scenario.

`r ref("Code snippet")`: Code to calculte the runoff prevention in each scenario.

@@

map_dfr(scenarios, runoff_prev) |> 
  mutate(scenario = names(scenarios), .before = 1) |> 
  kable()

@@

The total rainfall presented the same value in all scenarios because we use the same rain event and all scenarios represent the same total area. As urban agriculture becomes more present, the amount of rainwater harvested increases. However, runoff is not clearly influenced by urban agriculture because other green elements presents equal or even larger infiltration rates [@Clinton2018].

@@

# Conclusion

>>>>

```{r setup, include=FALSE}

# # This project uses {renv} for reproducibility. To install all dependencies in the proper version:
# renv::restore()

knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(ediblecity)
library(knitr)
library(sf)
library(purrr)
library(tidyr)
library(patchwork)
library(citr)

@@

Urban agriculture is gaining attraction as one of the pillars of the urban ecological transition and to increase food security in an urbanized planet. However, there is a lack of systematic quantification of the benefits provided by urban agriculture solutions. In this paper, we present an R package to estimate multiple benefits of urban agriculture. The goal is to provide a tool that help decision-makers to strategically implement urban agriculture solutions. It provides functions to calculate 8 indicators: urban heat island, runoff prevention, green areas accessibility, NO~2~ sequestration, jobs created in commercial gardens, volunteers involved in community gardens, green per capita and, finally, food production. Moreover, the package also provides a function to create new scenarios. We illustrate the use of the package by comparing three scenarios in a neighborhood of Girona (Spain). Using the data that is provided with the package as example data sets, we compare scenarios with an increasing amount of urban agriculture solutions. The `ediblecity` package is open-source software. This allows other R developers to contribute to the package providing new functionalities or improving the existing ones.

**Keywords:** edible city solutions; urban farming; urban food; nature-based solutions; Rstats; urban challenges; societal challenges; urban agriculture

@@

Therefore, decision-makers, who are responsible of leading the urban transitions to more sustainable and resilient cities, are orphan of evidence in how to implement urban agriculture to maximize its impact on sustainability. Yet, several studies provided some insights that can guide the implementation of urban agriculture. For instance, some models assist planners in rainwater harvesting [@Gittleman2017; @Lupia2017]. Broader, @GomezVillarino2021 developed guidelines to maximize ecosystem services through urban agriculture. And, as expected, many models have been developed to quantify food production and food security provided by urban agriculture [@Grafius2020; @Grewal2012; @Hsieh2017; @MacRae2010].

Hence, our goal is to provide a tool to estimate multiple benefits of urban agriculture that help decision-makers to strategically implement urban agriculture solutions. We developed the estimations for eight indicators and a functionality to create scenarios of urban agriculture based on the proportion of elements to be transformed to urban agriculture and on which elements will be transformed and packed all of them in an R package called `ediblecity`. In the methods, we present the main functionalities of the package and then we apply those to an example in a neighborhood of Girona (Spain).

@@

## Implementation

all the equations and algorithms to model the benefits of urban agriculture were encapsulated in an `R` package [@RCoreTeam2022], using one function for each indicator and one function to create new scenarios. The package was created using R version 4.2.1 [@RCoreTeam2022] in RStudio desktop v. 2022.07.02. The package structure was assisted by the package `devtools` [@devtools2022] and guided by @Wickham2022. Likewise, the documentation of the functions was assisted by the package `roxygen2` [@roxygen2022]. The dependencies of the package are:

- dplyr (>= 1.0.6) [@dplyr2022]
- magrittr (>= 2.0.1) [@magrittr2022]
- sf (>=0.9) [@sf2018]
- stars (>= 0.5) [@stars2022]
- rlang (>= 1.0) [@rlang2022]

### Urban representation of the city of interest

@@

city_functions |>
  select(-starts_with("...")) |> 
  select(-edible, -public, -jobs, -volunteers, -location) |>
  select(-starts_with("water")) |> 
  kable()
```
### Indicators estimated

#### Urban heat island

@@

#### Runoff prevention

@@

#### Green areas accessibility

@@

#### Nitrogen dioxide sequestration

@@

#### Jobs created and volunteers involved

@@

#### Green per capita

@@

#### Food production

@@

### Scenarios of urban agriculture

@@

## Operation

The `ediblecity` package is compatible with versions of R higher than 2.10. Although, it was created using version 4.2.1. `r ref("Code snippet", "install")` shows how to install the last development version, available in github.

`r cross_refs[["install"]]`: Code to install the last development version of the package

```{r, eval=FALSE, echo=TRUE}

# install.packages("devtools") # if not yet installed
devtools::install_github("icra/ediblecity")

Once the package is installed, it works as any R package. It can be attached to the namespace using `library(ediblecity)` or preceding the functions with `ediblecity::`. To check the documentation of the package and its functions, type `help(package="ediblecity")` in the R console.

## Limitations

As all abstraction of the reality, the equations and algorithms of the `ediblecity` package present some limitations. One limitation is the use of GIS layers to create scenarios and estimate indicators, which is intrinsically in two dimensions, and sometimes 2.5 dimensions, since we consider the height of buildings. This prevents to consider other urban agriculture solutions that are relevant such as vertical farming. To consider vertical solutions, the `ediblecity package` should include 3D calculations.

Another limitation is the subset of indicators chosen, this is, as we said, a choice. Other indicators might chosen instead or added to the current subset. Hopefully, the `ediblecity` package will be well received by the community of R scientists and other developers will add new indicators to fulfill their own needs. Indeed, this is one of the main advantages of open-source software.

@@

To use the `ediblecity` package in R, it is necessary to install it first. `r ref("Code snippet", "install")` shows how to install the last development version.

`r cross_refs[["install"]]`: Code to install the last development version of the package

```{r, eval=FALSE}

# install.packages("devtools") # if not yet installed
devtools::install_github("icra/ediblecity")

## Create scenarios of urban agriculture

To our understanding, the most important use case of the `ediblecity` package is to compare scenarios. To illustrate this, we will create two scenarios and compare them with the original `city_example`. The package has been designed to work well with the `tidyverse` framework in `R` [@Wickham2019], especially, with the `map_*` family in the `purrr` package [@Henry2022]. Therefore, we first create the two scenarios and save them in a list along with the original urban representation, which we called `s0` (`r ref("Code snippet", "use_set_scenario")`). The scenario `s1` will convert the 25% of elements to urban agriculture solutions while the scenario `s2` will convert the 100%. Half of gardens in vacant plots, streets and rooftop will have commercial purposes.

Since the scenarios and the indicators have some stochastic parameters, the ideal procedure would be to integrate the creation of the scenarios and the estimation of indicators in a Monte Carlo simulation to get the confidence intervals for each combination of scenario and indicator. However, we rather to keep things simple to better illustrate how to use the functions provided by the package.

@@

# Create new scenarios
scenarios <- map(c(0.25, 1), ~set_scenario(city_example, 
                             pGardens = .x,
                             pVacant = .x,
                             pRooftop = .x,
                             private_gardens_from = "Normal garden",
                             vacant_from = c("Vacant", "Streets"),
                             rooftop_from = "Rooftop",
                             pCommercial = 0.5))

@@

# Name the scenarios
names(scenarios) <- c("s1", "s2", "s0")
scenarios <- scenarios[order(names(scenarios))]
```

@@

counting(scenarios[[1]], "s0") |> 
  full_join(counting(scenarios[[2]], "s1")) |> 
  full_join(counting(scenarios[[3]], "s2")) |> 
  mutate(across(where(is.numeric), ~if_else(is.na(.x), 0L, .x))) |> 
  kable()
```

## Calculate indicators to compare scenarios

@@

# We use the SVF object that is provided for the city_example in the package
map_dfr(scenarios, UHI, SVF = ediblecity::SVF, .id = "Scenario") |> 
  kable()

@@

As expected the second scenario (`s2`) has the lowest values for urban heat island, but no too far from the scenario `s1`. Both present a reduction of approximated 50% in urban heat island regarding the base scenario. We can also generate a raster with the urban heat island for each scenario, like in `r ref("Figure", "UHI_plots")`

```{r, include=FALSE, eval = FALSE}

# plot <- map(scenarios, ~ UHI(.x, SVF = ediblecity::SVF, return_raster = TRUE), .id = "Scenarios") |> 
#   map(~ ggplot()+
#         stars::geom_stars(data = .x)+
#         coord_equal()+
#         scale_fill_distiller(palette = "RdYlBu")+
#         theme_void()+
#         theme(plot.title = element_text(hjust=0.5))
#       ) |> 
#   map2(names(scenarios), ~ .x+labs(title = .y)) |> 
#   wrap_plots()+plot_layout(guides = 'auto', nrow = 1)
# 
# ggsave("assets/UHI_plots.png")

![](assets/UHI_plots.png)
`r cross_refs[["UHI_plots"]]`: Raster returned by the `UHI` function when `return_raster` is set to `TRUE`.

`r ref("Code snippet", "runoff_table")`: Code to calculate the runoff prevention in each scenario.

@@

map_dfr(scenarios, runoff_prev, .id="scenario") |> 
  kable()

@@

The total rainfall presented the same value in all scenarios because we use the same rain event and all scenarios represent the same total area. Moreover, although there is an important reduction of runoff and an increase in rainwater harvested regarding the base scenario, there is no improvement from scenario 1 to scenario 2. As shown in `r cross_refs[["runoff_table"]]`, the rainwater harvested is larger in scenario 1. This is explained because the algorithm uses as catchment areas all adjacent upper areas that are not used for urban agriculture. Hence, as rooftop converted to urban agriculture increases, the availability of catchment areas decrease. The infiltration rates also increase because rooftop gardens retain water but this is not enough to compensate the reduction in harvesting.

@@

`r ref("Code snippet")`: Estimation of NO~2~ sequestration in all three scenarios

map_dfr(scenarios, no2_seq, .id="Scenario") |> 
  kable()
```

Regarding the capacity to absorb NO~2~, we see a significant improvement from base scenario to scenario 1 but not as larger from scenario 1 to scenario 2.

In the `r ref("Code snippet", "use_jobs")`, there is an example in job to calculate jobs as well as volunteers in the different scenarios.

`r cross_refs[["use_jobs"]]`: Code to estimate jobs and volunteers in each scenario.

jobs <- scenarios |> 
  map_dfc(edible_jobs, verbose = TRUE) |> 
  pivot_longer(everything(), values_to = "Jobs")

volunteers <- scenarios |> 
  map_dfc(edible_volunteers, verbose = TRUE) |>
  pivot_longer(everything(), values_to = "Volunteers")

bind_cols(jobs, volunteers["Volunteers"]) |> 
  pivot_longer(-name, names_to = "Indicator") |> 
  ggplot(aes(x=Indicator, y=value, fill=name))+
  geom_boxplot()+
  labs(fill="Scenario")

```
`r ref("Figure")`: Comparison of jobs created and volunteers involved in each scenario.

As expected, the base scenario presented very low values, it creates 0 jobs since there is not any commercial garden and a median of `r edible_volunteers(scenarios[[1]])[[2]]` volunteers, involved in one community garden. The number of jobs and volunteers also increases from scenario 1 to scenario 2 as well as the uncertainty related to the numbers.

Below, we calculate the green per capita in each neighborhoods. To do so, we use another spatial data set provided by the `ediblecity` package (`neighbourhoods_example`) which contains the neighborhoods of `city_example` along with the inhabitants in each neighborhood.

`r ref("Code snippet")`: Code to calculate the green per capita in each neighborhood

scenarios |> 
  map_dfr(green_capita, 
          neighbourhoods = neighbourhoods_example, 
          inh_col = "inhabitants",
          name_col = "name",
          private = TRUE,
          verbose = TRUE,
          .id = "scenarios") |> 
  ggplot(aes(x=name, y=green_capita, fill=scenarios))+
  geom_col(position = position_dodge())+
  labs(x="Neighborhoods", y=bquote(m^2/person), fill="Scenario")

```
`r ref("Figure")`: Comparison of green per capita in each neighbourhood across scenarios.

The differences between both neighborhoods is due to their urban origin. Sant Narcís nord was designed like a city garden while Sant Narcís sud is mainly composed of apartments. The interesting issue is that the improvement across scenarios is larger in Sant Narcís nord that in Sant Narcís sud, evidencing that an increase in urban agriculture is not enough to achieve environmental justice unless it is ideologically planned [@Jennings2012].

The last, but not least, indicator provided by the `ediblecity` package is the food production. The food production is assumed as higher in gardens for commercial purposes than in community gardens, which the goal is not to maximize the production, especially in the case of rooftops, where the commercial rooftop gardens are assumed to use hydroponic technology while community rooftop gardens are assumed to use raised beds following the study of @Caputo2020.

`r ref("Code snippet")`: Code to get confidence intervals of food production in Tm/year.

scenarios |> 
  map_dfr(food_production, .id = "Scenario") |> 
  mutate(across(where(is.numeric), ~ .x/1000)) |> 
  kable()

Although the median are clearly different, we cannot state that food production is bigger in scenario 2 with a 95% of confidence (i.e. p-value > 0.05 in differences between `s1` and `s2`). Taking the most optimistic scenario (`s2` at quantile 95%), the urban agriculture in our example could produce `r prod = food_production(scenarios[[3]])[[3]]/sum(neighbourhoods_example$inhabitants); prod` kg/year/person. The daily intake of fruits and vegetables recommended by FAO is 200 gr/person, i.e. 73 kg/person/year [@FAO2004]. Therefore, our optimistic estimation would provide `r prod/73` times the neighborhood's needs in fruits and vegetables. However, it would require (taking also the higher interval) `r jobs = edible_jobs(scenarios[[3]])[[3]];round(jobs)` people working in commercial gardens and `r vol = edible_volunteers(scenarios[[3]])[[3]];round(vol)` volunteers involved in community gardens, which is `r (jobs + vol)/sum(neighbourhoods_example$inhabitants)` times the inhabitants of the neighborhood.

# Conclusions

In this paper, we presented the `ediblecity` package. An R package to to model and estimate the benefits of urban agriculture. The package is ready to be used by R users with a basic level. It can be used to estimate the benefits of real cases as well as to create scenarios. In both cases, 8 indicators can be calculated. In the example illustrated in this paper, there is a lot of uncertainty because the stochastic parameters are randomized using a large range. However, all these parameters can be customized by the user in case they have more accurate data for the case study at stake. With more accurate data, the uncertainty can be easily reduced changing the arguments of the functions. Likewise, some assumptions of the models can be override with truly statements.

The `ediblecity` package is open-source software under MIT license. This allows other R developers to contribute to the package providing new functionalities or improving the existing ones. Therefore, an open-source software is always a work-in-progress.

## Data Avaialability statement {.unnumbered}

All data underlying the results are available as part of the `ediblecity` package and no additional source data are required.

## Software availability {.unnumbered}

- The `ediblecity` package is available at [https://github.com/icra/ediblecity](https://github.com/icra/ediblecity) under MIT license.
- This paper has been entirely written using RMarkdown, the code to reproduce the paper is available at [https://github.com/icra/ediblecity_paper](https://github.com/icra/ediblecity_paper) under MIT license.

## Authors contributions {.unnumbered}

JPR, JC and LC contributed to conceptualization, methodology and writing - review & editing. JPR contributed to data curation, formal analysis, software, validation, visualization and writing - original draft preparation. JC and LC contributed to funding acquisition, project administration, resources and supervision.

## Competing interests {.unnumbered}

No competing interests were disclosed.

## Grant information {.unnumbered}

Authors acknowledge the support from the European Union Horizon 2020 innovation & action project EdiCitNet (GA776665), the Economy and Knowledge Department of the Catalan Government through Consolidated Research Groups ICRA-ENV 2017 SGR 1124 and 2017 SGR 1318, and the CERCA program.

## Acknowledgements

`r stop("add acknowledgements to R commnunity")`
